---
title: "Analysis of Vehicle Collisions in Canada"
output:
  html_document:
    toc: true
    toc_depth: 1
    theme: united
---

Analysis of Vehicle Collisions in Canada 
========================================================
### Jas Sohi 
#### April 1, 2015

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.
# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

###Set global options for all subsequent r code chunks 
library(knitr)
opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(ggplot2)
library(plyr)
#http://rstudio.github.io/DT/
library(DT) #new package that allows R data objects to be displayed as tables 
#on HTML pages
#Global option to have a black and white theme for all plots
ggplot <- function(...) { ggplot2::ggplot(...) + theme_bw() }

#http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)
#multiplot function to plot multiple graphs on a single page

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist 
# (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
#http://stackoverflow.com/questions/19233365/how-to-create-a-marimekko-mosaic-
#plot-in-ggplot2
#Mosaic Plots in ggplot2
ggMMplot <- function(var1, var2, xlabel, ylabel, title){
  require(ggplot2)
  levVar1 <- length(levels(var1))
  levVar2 <- length(levels(var2))

  jointTable <- prop.table(table(var1, var2))
  plotData <- as.data.frame(jointTable)
  plotData$marginVar1 <- prop.table(table(var1))
  plotData$var2Height <- plotData$Freq / plotData$marginVar1
  plotData$var1Center <- c(0, cumsum(plotData$marginVar1)[1:levVar1 -1]) +
    plotData$marginVar1 / 2

  ggplot(plotData, aes(var1Center, var2Height)) + 
    geom_bar(stat = "identity", aes(width = marginVar1, fill = var2), 
             col = "Black") +
    geom_text(aes(label = as.character(var1), x = var1Center, y = 1.05)) +
    xlab(xlabel) + #added these three lines - Jas
    ylab(ylabel) +
    ggtitle(title)
  }

```

### About the Dataset

[National Collision Database (NCDB) - a database containing all police-reported motor vehicle collisions on public roads in Canada. Selected variables (data elements) relating to fatal and injury collisions for the collisions from 1999 to the most recent available data (2011).](http://open.canada.ca/data/en/dataset/1eb9eba7-71d1-4b30-9fb1-30cbdab7e63a)

### Goal

####*The underlying question I will explore in this analysis is what features (related to drivers, their vehicles, and the collision itself) are common in accidents that result in fatalities?*

### Data Wrangling

I chose to subset this data and only look at the most recent year (2011) due to the enormous size of the original dataset with over a million records. This new raw dataset contains just over 300,000 records.

Each record represents one collision report per person involved in a collision in Canada. So if there were 4 people involved in a collision there would be 4 separate records. 

```{r}
###DATA PREPARATION - This was done only once to subset the original da
###Set the working directory
# getwd()
# directory <- "C:/Users/user/version-control/Project-3-Data-Analysis-with-R"
# setwd(directory)
###Load up the original dataset
# df <- read.csv('NCDB_1999_to_2011.csv')
###Subset the dataset
# df2011 <- subset(df, C_YEAR == "2011")
# write.csv(df2011, "NCDB_2011.csv")
# rm(list=ls()) #removes all variables from the Global Environment 
```

```{r}
# Load the Data
filename <- "NCDB_2011.csv"
df <- read.csv(filename)
df2 <- df #Copy of the untouched dataframe for later comparison
```

### Number of records, variables 
```{r}
nrow(df) 
ncol(df)
```
### Variable names
```{r}
names(df)
```
[Data Dictionary (.doc)](http://data.tc.gc.ca/extracts/Data_Dictionary_NCDB_EN.doc)

# Univariate Analysis

### What is the structure of your dataset?

### Structure of the data
```{r}
str(df)
```

### Summary of the data
```{r}
summary(df)
```

There are three kinds of variables in this dataset: collision-level, vehicle-level, and person-level.

### COLLISIONS

From an initial look a the summary output, Friday (5) appears to be the weekday with the most number of collisions reported. The 4-5 PM Hour (16) is the most dangerous time for motorists. Most collisions involve 2 vehicles (*C_VEHS*). Rear-end collisions (21) are the most common types of accidents (*C_CONF*) and most accidents happen at an intersection (02) (*C_RCFG*). Nothing out of the ordinary so far. 
However, perhaps surprisingly, most accidents actually occur on clear and sunny days by a wide margin (1) (*C_WTHR*) and on dry,
normal (1) (*C_RSUR*) straight and level (1) 
(*C_RALN*) road surfaces. Finally, most collisions occur where no traffic controls (such as stop signs, police officers, reduced speed zone, etc.) are present.

```{r}
datatable(head(df[,2:13]), class = 'cell-border stripe', rownames = FALSE, 
          caption = 'Table 1: Collision Level Variables (head)', 
          options = list(dom = 't'))
```

### VEHICLES

Most collisions involved light duty vehicles such as passenger cars, vans, and light duty pickup trucks (01) (*V_TYPE*).
Most collisions involve more recent model vehicles, but there isn't a large variance between years (*V_YEAR*). 

```{r}
datatable(head(df[,14:17]), class = 'cell-border stripe', rownames = FALSE,
          caption = 'Table 2: Vehicle Level Variables (head)', 
          options = list(dom = 't'))
```

### PEOPLE

```{r echo = FALSE}
#Divide the number of deadly collisions by the number of total collisions
percent_deadly <- round((nrow(subset(df,P_ISEV == 3))/nrow(df))*100,digits=2)
```

Men are involved in more collisions than women. Many people don't report their age (UU = Unknown) (*P_AGE*), but from the ones that do, most consist of young adults and teenagers (17-21 years old). About 2/3rds of the reported collisions (each record) are from drivers (11) themselves instead of passengers (*P_PSN*). Only ```r percent_deadly``` % of reported collisions involved a fatality (3), while the rest of the collisions were split mostly between injuries/no injuries (*P_ISEV*). In the vast majority of incidents, a safety device such as a seat belt or helmet (for motorcycles, bicyclists,
snowmobile, and ATV riders) was worn (02) (*P_SAFE*).
This highlights the fact that there are different categories of people reporting these collisions, but the vast majority are vehicle drivers (1) and passengers (2) with pedestrians (3) a distant third (*P_USER*).  

```{r}
datatable(head(df[,17:23]), class = 'cell-border stripe', rownames = FALSE,
          caption = 'Table 3: People Level Variables (head)', 
          options = list(dom = 't'))
```

# Univariate Plots Section

###Initial Age Plot
```{r}
qplot(x=P_AGE, data = df)
```

Looks like there is something unusual with this plot towards the right tail.

```{r}
#Let's see the frequency of each age
ages <- as.data.frame(table(df$P_AGE))
ages[ages$Var1 == "NN" | ages$Var1 == "UU",]
summary(df$P_AGE)[1:10]
```

### Further data wrangling

Looks like a lot of people did not report their age (UU) or was not applicable (NN) (e.g. “dummy" person record created for parked cars).
We'll go ahead and remove these records from the dataset and also clean up the gender variable to show only males and females. 

```{r}
df <- subset(df, P_AGE != "UU" & P_AGE != "NN" & P_SEX != "N" & P_SEX != "U")
#summary(df)
#275,270 records now
```

We are now looking at `r nrow(df)` records instead of just over three hundred thousand from the original data. We got rid of under 10% of the records and it shouldn't affect our ultimate analysis.

### Second Age Plot
```{r}
ggplot(aes(x=P_AGE), data = df) + geom_bar()
```

This chart now shows the distribution of ALL people involved in collisions and we can see that it is bimodal with one peak near later teenagers/young adults and another what looks like people in their 40s. Let's confirm that.

```{r}
#Needed to subset df & for later boxplot.
df$P_AGE_NUM <- as.numeric(df$P_AGE)
ggplot(aes(x=P_AGE), data = subset(df,P_AGE_NUM > 15)) + 
  geom_bar(binwidth = 1)
```

From this chart (age > 15), we can see more precisely that 45-50 is where the second peak is. But, as per the original goal of this analysis, we are really interested in the age of *Drivers* only.

### Driver's Age

```{r}
#P_USER is the type of user who reports the collision, 
#can be passenger, pedestrian, motorcyclists, etc.
ggplot(aes(x=P_AGE), data = subset(df,P_USER == 1)) + #1 for Driver
  geom_bar(binwidth = 1)
```

We do see a similar distribution when only considering Drivers only. We now see that the left tail of the distribution seems flat as we don't expect mqny children to be driving (at least in Canada!)

###Gender of the Driver

Are more men or women behind the wheel before an accident?

```{r}
#http://stackoverflow.com/questions/20600900/r-faceted-bar-chart-with-
#percentages-labels-independent-for-each-plot
ggplot(aes(x=P_SEX),data = subset(df,P_USER == 1)) + geom_bar() + 
  stat_bin(geom = "text", #show percentages on top of the bars
           aes(label = paste(round((..count..)/sum(..count..)*100), "%")),
           vjust = -1, color = "grey30", size = 5) +
           coord_cartesian(ylim=c(0,125000)) #leave room for the percent labels
```

[Contrary to this reddit joke](https://www.reddit.com/r/Jokes/comments/2zqnwx/fact_a_lot_of_women_turn_into_good_drivers/) (in bad taste I may add) men are actually involved in more collisions than women overall.

### Deadly Crashes

Is this difference between the sexes still present when we look at deadly crashes only?

```{r}
#C_SEV Severity of the crash - At least one fatality(1) or non-fatal injuries(2) 
ggplot(aes(x=P_SEX),data = subset(df,P_USER == 1 & C_SEV == 1)) + geom_bar() +  
  stat_bin(geom = "text", #show percentages on top of the bars
           aes(label = paste(round((..count..)/sum(..count..)*100), "%")),
           vjust = -1, color = "grey30", size = 5) +
           coord_cartesian(ylim=c(0,2250)) #leave room for the percent labels
```

Yes, and the contrast is much more pronounced. There are relatively a few deaths per year (thank god!), but in 3 out of every 4 collisions that do involve a death, a man was driving the vehicle.

### Time

Let's shift gears (no pun intended) and look at collisions in regards to time.

### What day of the week has the most accidents?

```{r}
ggplot(aes(x=C_WDAY), data = subset(df,C_WDAY != "U")) + geom_bar()
```

Looks like Friday (5) has the most accidents (an increasing trend from Monday - perhaps people are in a rush to celebrate the weekend!) and Sunday has the least amount of accidents (this could be attributed to the fact that very few people work on Sundays so there might be less vehicles commuting to work overall).

```{r}
g1 <- ggplot(aes(x=C_WDAY), data = subset(df,C_WDAY != "U" & C_SEV == 1)) + 
  geom_bar() + ggtitle("Deadly Collisions")
g2 <- ggplot(aes(x=C_WDAY), data = subset(df,C_WDAY != "U" & C_SEV == 2)) + 
  geom_bar() + ggtitle ("Non-Deadly Collsions")
multiplot(g1, g2, cols=2)
```

Looks like deadly collisions happen the most on Saturdays (6) as opposed to Fridays (5) for non-deadly crashes. 

### What time of day do most collisions occur?

```{r}
ggplot(aes(x=C_HOUR), data = df) + geom_bar()
```

Not surprisingly, most collisions occur during 4:00-5:00 PM (not surprisingly many people are heading home after work during this hour).

```{r}
g1 <- ggplot(aes(x=C_HOUR), data = subset(df,C_SEV == 1)) + geom_bar() + 
  ggtitle("Deadly Collisions")
g2 <- ggplot(aes(x=C_HOUR), data = subset(df,C_SEV == 2)) + geom_bar() + 
  ggtitle ("Non-Deadly Collsions")
multiplot(g1, g2, cols=2)
```

For deadly collisions, a larger proportion occur during the sleeping hours (from midnight to before the start of the workday).

### What month has the most accidents?

```{r}
ggplot(aes(x=C_MNTH), data = subset(df, C_MNTH != "UU")) + geom_bar()
```

Looks like January (01) and the summer months (07 & 08) have the most number of reported collisions (could be that more people are driving in the summer for the holidays).

### Does this change when we only look at drivers?

```{r}
ggplot(aes(x=C_MNTH), data = subset(df, C_MNTH != "UU" & P_USER == 1)) + 
  geom_bar()
```

Not really, the absolute counts are less, but the distribution is quite similar.

### Types of People Reporting Collisions

Speaking of drivers, we haven't yet discussed how much of this dataset consists of collision reports from drivers as opposed to passengers, pedestrians, etc.

```{r}
ggplot(aes(x=P_USER), data = subset(df, P_USER != "U")) + geom_bar() +
  stat_bin(geom = "text", #show percentages on top of the bars
           aes(label = paste(round((..count..)/sum(..count..)*100), "%")),
           vjust = -1, color = "grey30", size = 4) +
           coord_cartesian(ylim=c(0,200000)) #leave room for the percent labels
```

The vast majority of this dataset consists of reports from drivers (1) and passengers (2).

### How's the weather like when there are accidents?

```{r}
ggplot(aes(x=C_WTHR), data = df) + geom_bar()
```

Perhaps counterintuitively, most accidents occur on sunny days (1). At first, you might have thought that more accidents occur in bad weather.

### Weather for motorcyclists vs drivers

```{r}
g1 <- ggplot(aes(x=C_WTHR), data = subset(df, P_USER == 5)) + #5 - Motorcyclists
  geom_bar() +
  ggtitle("Motorcyclists")
  #More accidents in rain and snow
g2 <- ggplot(aes(x=C_WTHR), data = subset(df, P_USER == 1)) + #1 - Drivers
  geom_bar() +
   ggtitle("Drivers")
multiplot(g1, g2, cols=2)  
 
```

A larger proportion of the accidents that motorcyclists are involved in occur on sunny days (1) than compared with motor vehicle drivers (this could simply be due to the fact that a lot of motorcyclists ride only in the spring and summer months with better weather conditions). 

### Number of vehicles involved in a collision

```{r}
qplot(x=C_VEHS, data = subset(df, P_USER == 1)) 
```

Most collisions involve, not surprisingly, 2 vehicles.

### Vehicle Age

```{r}
qplot(x=V_YEAR, data = subset(df, P_USER == 1)) 
```

```{r}
g1 <- ggplot(aes(x=V_YEAR), data = subset(df, P_USER == 1 & C_SEV == 1)) +
  geom_bar() +
  ggtitle("Deadly Collisions")
g2 <- ggplot(aes(x=V_YEAR), data = subset(df, P_USER == 1 & C_SEV == 2)) +
  geom_bar() +
  ggtitle("Non-Deadly Collisions")
multiplot(g1,g2, ncol = 2)
```

Let's cleanup the plot a bit and only look at vehicles manufactured after 1980.

```{r}
#Need to create a new numeric variable to be able to subset the data below
df$V_YEAR_NUM <- as.numeric(as.character(df$V_YEAR))
g1 <- ggplot(aes(x=V_YEAR), data = subset(df, P_USER == 1 & C_SEV == 1 &
  V_YEAR != "NNNN" & V_YEAR != "UUUU" & df$V_YEAR_NUM >= 1980)) +
  geom_bar() +
  ggtitle("Deadly Collisions")
g2 <- ggplot(aes(x=V_YEAR), data = subset(df, P_USER == 1 & C_SEV == 2 & 
  V_YEAR_NUM >= 1980 & V_YEAR != "NNNN" & V_YEAR != "UUUU")) +
  geom_bar() +
  ggtitle("Non-Deadly Collisions")
multiplot(g1,g2, ncol = 2)
```

```{r}
qplot(geom = "boxplot", x= factor(C_SEV), y = V_YEAR_NUM, data = df)
```

Doesn't seem to be any clear trend other than most vehicles involved in accidents are from most recent models (it could simply be there are more newer vehicles on the road). Looks like a dead end here (pun intended!).

###Vehicle Type

Okay, so now the type of vehicle you are driving has to be a factor in determining whether deaths are involved in collisions right? Let's see.

```{r echo=FALSE, Summary}
qplot(x=V_TYPE, data = df)
```

Looks like the vast majority of vehicles involved in accidents are Light Duty Vehicles (1).

```{r}
g1 <- ggplot(aes(x=V_TYPE), data = subset(df,P_USER == 1 & C_SEV == 1)) +
  geom_bar() +
  ggtitle("Deadly Collisions")
g2 <- ggplot(aes(x=V_TYPE), data = subset(df,P_USER == 1 & C_SEV == 2)) +
  geom_bar() +
  ggtitle("Non-Deadly Collisions")
multiplot(g1,g2, ncol = 2)
```

Now remember we are now looking at drivers only; we see that a larger proportion of Trucks (07) & Road Tractors (08) are involved in deadly crashes. This may be one variable to help predict deadly crashes.

#Univariate Analysis (continued)

### What is/are the main feature(s) of interest in your dataset?

The main features of interest in the data set are *C_SEV* (the severity of collisions - injury, no injury, or fatality) and *P_AGE* (the age of person involved in collision). I suspected that driver's age and some other combination of variables could help predict deadly collisions.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

In addition to the age of the driver, I originally thought that vehicle age (*V_YEAR)*, vehicle type (*V_TYPE*), and gender (*P_SEX*) would be a larger factor in determining the fatality in collisions. The plots support this initial hunch except for vehicle age which doesn't seem to vary between deadly and non-deadly collisions. I will look further at age in the bivariate analysis to see if their is a difference between deadly and non-deadly collisions.

### Did you create any new variables from existing variables in the dataset?

Yes, I did create new variables just so that the data could be subset since the original variables were factors (categorical) and could not be subsetted otherwise (*P_AGE_NUM* & *V_YEAR_NUM*). I didn't want to directly convert the original columns in case I needed the factor variables at a later time.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

Yes, I did adjust the form of the data to show only the most recent year of data as there were millions of total records in the initial dataset. As mentioned earlier, I subset 300,000 records and then I removed data that had unknown or missing gender values as I suspected that the sex of the driver in fatal collisions could be skewed towards one sex (as confirmed by a histogram plot).

#Bivariate Analysis

### Further Data Wrangling 

Before we get to bivariate plots, let's do some additional data wrangling.

```{r}
#http://stackoverflow.com/questions/3571909/r-how-to-calculate-correlation-cor-
#for-only-a-subset-of-columns
# Make the correlation table
remove <- names(df) %in% c("C_YEAR") #YEAR produces NA values for this table
rdf <- df[!remove]
ctab <- cor(rdf[sapply(rdf, is.numeric)])
round(ctab, 2)
```

This initial correlation table with the current numeric variables doesn't show us much valuable info.
Let's see what other variables can be considered numeric so we can create a more useful correlation table.

```{r echo=TRUE}
df$C_MNTH_NUM <- as.numeric(df$C_MNTH); 
df$C_WDAY_NUM <- as.numeric(df$C_WDAY)
df$C_HOUR_NUM <- as.numeric(df$C_HOUR); 
df$C_VEHS_NUM <- as.numeric(df$C_VEHS) #number of vehicles
df$C_WTHR_NUM <- as.numeric(df$C_WTHR) #higher value = worse weather
df$C_RSUR_NUM <- as.numeric(df$C_RSUR) #higher value = worse road conditions
#Create a new variable, M (Mortality), 1 = Died, 0 = Survived 
df$M <- ifelse(df$C_SEV == 1, 1, 0) 
#However, some 0s are unknown values that need to be removed
```

We have created 6 new numeric variables from the original factor variable and new binary variable called *M* (Mortality). *1* means at least one person died immediately after the collision or went to the hospital and died as a result of his/her injuries. *0* is everyone involved survived the crash. However, this new *M* variable is based off of the *C_SEV* variable we looked at and there may be some unknown variables that still need to remove.

```{r}
#275,270 records before
df3 <- df  #create a copy in case need to revert due to unintended changes
#U - known, X - jurisdiction does not provide info - is not found in this 2011
df <- subset(df, C_SEV != "U" | C_SEV != "X") 
```

There happen to be no such unknown records for 2011. We will also further subset the data to only look at drivers since that's what our original question asked: what features (*related to drivers*, their vehicles, and the collision itself) are common in accidents that result in fatalities?

```{r}
df4 <- df
#275,270 records before
df <- subset(df, P_USER == 1)
```

Now we have reduced the records by just under 100,000 (`r nrow(df)` records) and we can look at the correlation matrix now.

#Bivariate Plots

```{r}
#Only look at the numeric variables in the dataframe since correlation doesn't
#make sense for factors
ctab <- cor(df[sapply(df, is.numeric)])
round(ctab, 2)
```

Definitely, there doesn't seem to be any linear relationships here. Except for a decent correlation between Weather & Road Surface (*C_WTHR_NUM* & *C_RSUR_NUM*) of `r round(cor(df$C_WTHR_NUM,df$C_RSUR_NUM),2)` 

```{r}
qplot(C_RSUR_NUM,C_WTHR_NUM, data = df, geom="jitter",  malpha = 0.5)
```

No clear relationship here. Looks like there are quite a few accidents when the weather is really bad and when the road surface is at its worst (top right hand corner).

Let's see if we can get any inspiration to compare other bivariate relationships with a pairwise plot of some selected variables.

```{r}
#P_PSN,P_USER don't need to be looked at as they both indicate drivers only
remove_columns <- names(df) %in% 
  c("X","C_YEAR","C_MNTH", "C_HOUR", "V_ID","V_YEAR","P_ID", "P_AGE", "P_ISEV",
    "C_SEV", "P_USER", "C_WTHR", "C_VEHS", "C_CONF", "C_RCFG","C_TRAF","P_SAFE",
    "P_PSN", "C_WDAY") 
sampledf <- df[!remove_columns]
#Rename the variables for better visibility on the plot
names(sampledf) <- c("RS","RL","VT","SEX","AGE","VY","Mon","D","H","VN","WT",
                     "M")
#https://tgmstat.wordpress.com/2013/11/13/plot-matrix-with-the-r-package-ggally/
library(GGally)
ggpairs(data=sampledf[1:10000, c(3:12)], # data.frame with variables,
        title="pairwise plots") # title of the plot
```

Let's see what other relationships we can explore further.

### Drivers in Newer Vehicles or Older Vehicles?

```{r}
ggplot(aes(x=factor(M), y=V_YEAR_NUM), data=subset(df,V_YEAR != "UUUU" & 
      V_YEAR != "NNNN")) + geom_boxplot()
```

Looks like newer cars are involved in most accidents whether deadly (1) or not (0).

### Vehicle Types


```{r}
#Using the mosaic plot function defined in first code chunk
g1 <- ggMMplot(df$V_TYPE[df$M==1],df$P_SAFE[df$M==1], "Vehicle Type", 
               "Safety Device Used","Deadly Collisions")
g2 <- ggMMplot(df$V_TYPE[df$M==0],df$P_SAFE[df$M==0], "Vehicle Type", 
               "Safety Device Used", "Non-Deadly Collisions")
```

```{r}
g1
```

```{r}
g2
```

Looks like many deadly crashes had no safety device such as a seat belt used (light pink) (this is consistent across the major types of vehicles).

Let's look at the age distribution of drivers of different types of vehicles.

```{r, out.width='1000px', dpi=200}
vehicle_types <- c("CAR\n(1)","VAN\n(5)","SMALL TRUCK\n(6)","LARGE TRUCK\n(7)",
                   "TRACTOR\n(8)","SCHOOLBUS\n(9)", "SCHOOLBUS\n(SMALL)\n(10)", 
                   "CITY BUS\n(11)", "RV\n(18)", "FIRETRUCK\n(21)",
                   "STREET CAR\n(23)")
ggplot(aes(x=V_TYPE,y=P_AGE_NUM), data = df) + 
  geom_boxplot() +
  scale_x_discrete(labels=vehicle_types) 
```

Nothing out of the ordinary, we can see that school bus drivers (9) and RV drivers (18) skew towards older drivers.

Let's now explore what traffic controls appear to be more effective than others.

```{r}
qplot(x=C_TRAF, geom="bar", fill=factor(M), data = df)
```

```{r}
ggplot(df, aes(x = C_TRAF, fill = factor(M))) + geom_bar(position = "fill")
```

From the top plot it appears as though no controls (18) being present would be the predominant traffic control when deadly accidents occur. But looking at the proportions, it looks as though the largest proportion of deaths actually occur where warning signs (5) are present.

Once again we have to remember that we are looking at the difference between deadly and non-deadly crashes.

### Age of Deadly Drivers
```{r}
library(vcd)
#http://cran.us.r-project.org/web/packages/vcdExtra/vignettes/vcd-tutorial.pdf
cdplot(factor(M) ~ P_AGE, data = df)
#with(Arthritis, rug(jitter(Age), col="white", quiet=TRUE))
```

This cumulative density plot shows teenage drivers and elderly drivers are overrepresented in collision deaths.

### Maybe the configuration of the crash and the road has an effect on how deadly a crash is?

```{r}
ggplot(aes(x=C_CONF, fill= factor(M)), data = df) + geom_bar(position = "fill")
```

Head on collisions (31)  make up a huge proportion of deadly crashes. 

```{r}
ggplot(aes(x=C_RCFG, fill= factor(M)), data = df) + geom_bar(position = "fill")
```

Railroad crossings (4) and passing lanes (7)  are the type of road configurations
with the largest proportion of deadly crashes.

### Time Revisted - How about if we look at time of day across the different days?

```{r}
ggplot(aes(x=C_HOUR), data = subset(df,C_WDAY != "U" & C_HOUR != "UU")) + 
  geom_bar() +
  facet_wrap(~C_WDAY)
```

Looks like distribution across the weekdays is bimodal throughout the "non-sleeping" hours (7 AM to Midnight), while the distribution on the weekend is almost identical in shape for Saturday & Sunday (being nearly unimodal for this same hour range).

### Gender

Does age distribution vary across genders?

```{r}
ggplot(aes(x=P_AGE), data = subset(df, P_USER == 1)) + geom_bar() +
  facet_wrap(~P_SEX)
qplot(x = P_SEX, y = P_AGE_NUM, data = df, geom = "boxplot")
```

Not quite, the shape of both age distributions is very similar across sexes.

# Bivariate Analysis (continued)

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

Railroad crossings (4) and passing lanes (7)  are the type of road configurations
with the largest proportion of deadly crashes. Head on collisions (31)  make up a huge proportion of deadly crashes. Interestingly, warning signs represent the traffic control category that has the largest proportion of deadly crashes.

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

Yes, there was an interesting relationship between hour and weekday. Looks like distribution across the weekdays is bimodal throughout the non-sleeping hours (7 AM-Midnight), while the distribution on the weekend is almost identical in shape for Saturday & Sunday (being nearly unimodal for this same hour range). 

### What was the strongest relationship you found?

The strongest relationship I found was between Weather (*C_WTH_NUM*) and Road Surface (*CRSUR_NUM*). There was a correlation of 0.43 between these two features. However, when plotting these two discrete values, no interesting trend was observed.

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
qplot(x=P_AGE, y=V_YEAR, data=subset(df,V_YEAR != "UUUU" & V_YEAR != "NNNN"), 
color = P_SEX)
```

No clear trend can be observed when plotting vehicle year, person's age, and sex except what was seen before in an earlier plot that the vast majority of vehicles involved in collisions are newer vehicles.

```{r}
qplot(x=V_TYPE, y=P_AGE_NUM,  data=subset(df,V_YEAR != "UUUU" & 
      V_YEAR != "NNNN"), color = P_SEX, geom="jitter")
```

It's interesting to see that the age distribution of school bus drivers (09) and city bus drivers (11) is about the same (about 25 to 75). However, school bus drivers have about the same number of women and men involved in accidents (women slightly more), while city bus drivers have predominately more males in accidents.

```{r}
qplot(x=C_MNTH, y=factor(M),  data=df, color = V_TYPE , geom="jitter")
```

Another dead end, no visible relatioship between month, vehicle type, and mortality (deadly or non-deadly crashes).

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

The only real observation that was observed was that the age distribution of school bus drivers (09) and city bus drivers (11) is about the same (about 25 to 75). However, school bus drivers have about the same number of women and men involved in accidents (women slightly more), while city bus drivers have predominately more males in accidents.

### Were there any interesting or surprising interactions between features?

Nothing surprising, but I did expect to see more multivariate relationships than what I found.

------


# Final Plots and Summary

### Plot One
```{r echo = FALSE, eval=FALSE}
# To help see what limits to set on next plot
boxplot(df$P_AGE_NUM) 
summary(df$P_AGE_NUM)
```

```{r, Plot_One, echo =FALSE}
#http://stackoverflow.com/questions/23709060/change-r-markdown-plot-width
#how may deaths for the most deadly age (19) and the second peak
age_most <- 19
age_old <- 46 #2nd peak
age_initial <- 14 #lowest age on the bar chart
most_deaths <- length(which(df$P_AGE_NUM == age_most))
older_deaths <- length(which(df$P_AGE_NUM == age_old))
#I will choose 14 as the minimum since that is the lowest legal age
g1 <- ggplot(aes(x=P_AGE), data = subset(df,P_AGE_NUM >= age_initial & 
  P_AGE_NUM <= 90)) + 
  geom_histogram(binwidth = 1, colour = "darkgreen", aes(fill = P_SEX)) +
  ggtitle("Histogram of Driver's Age (All Collisions)") + 
  xlab("Driver's Age (P_AGE)") + 
  ylab("Count of Drivers") +
  #http://stackoverflow.com/questions/11019692/place-annotation-at-the-top-of-a
  #-series-of-histograms-in-ggplot2-using-a-for-loo
  annotate("text", label = paste("Age 19: ",as.character(most_deaths)), 
    x= age_most - age_initial, hjust = 0, y = most_deaths + 100,
    color= "black", size = 4) + #Show the exact number
  annotate("text", label = paste("Age 46: ",as.character(older_deaths)), 
    x= age_old - age_initial, hjust = 0, y = older_deaths + 100, 
    color = "black", size = 4) + #Show the exact number
   #added 100 to the y-axis so there would be no overlap between text and plot
  theme(legend.title=element_blank(), legend.position = "top", 
         plot.title = element_text(lineheight=.8, face="bold")) +
  scale_x_discrete(breaks = seq(15,90,5), labels = seq(15,90,5) ) +
  scale_fill_manual(values=c("pink", "lightblue"))
#2nd graph
deadliest_age <- 18 #can be seen from visually inspecting the graph 
deadly_count <- length(which(df$P_AGE_NUM == 18 & df$M == 1)) 
g2 <- ggplot(aes(x=P_AGE), data = subset(df,P_AGE_NUM >= age_initial & 
  P_AGE_NUM <= 90 & M == 1)) + 
  geom_histogram(binwidth = 1, colour = "darkgreen", aes(fill = P_SEX)) +
  ggtitle("Histogram of Driver's Age (Deadly Collisions)") + 
  xlab("Driver's Age (P_AGE)") + 
  ylab("Count of Drivers") +
  scale_y_discrete(breaks = seq(0,85,5), labels = seq(0,85,5)) +
  scale_x_discrete(breaks = seq(15,90,5), labels = seq(15,90,5)) + 
  #http://sape.inf.usi.ch/quick-reference/ggplot2/colour - color names
  scale_fill_manual(values=c("pink", "lightblue")) +
  theme(legend.title=element_blank(), legend.position = "top", 
    plot.title = element_text(lineheight=.8, face="bold")) +
  annotate("text", label = "*", x=5.5 , hjust=0, y=deadly_count - 5, color= "black",
    size=20)#show an asterisk
```

```{r, out.width='1000px', dpi=200}
g1
```

```{r, out.width='1000px', dpi=200}
g2
```

#### From the top plot we can see that the distribution of driver's age is bimodal (1st peak at age 19 and 2nd peak at age 46). 19 year olds are indeed involved in the most accidents overall, but 18 year olds (*) are the ones involved in most deadly collisions (lower plot). We can also see that men and women are about equal in terms of being involved in accidents. However, when we look only at deadly collisions only (lower plot) we see that men are overrepresented. 


### Plot Two
```{r echo=FALSE,out.width='1000px', dpi=200, Plot_Two}
#Customized Mosaic Plot function
ggMMplot <- function(var1, var2, xlabel, ylabel, title){
  require(ggplot2)
  levVar1 <- length(levels(var1))
  levVar2 <- length(levels(var2))

  jointTable <- prop.table(table(var1, var2))
  plotData <- as.data.frame(jointTable)
  plotData$marginVar1 <- prop.table(table(var1))
  plotData$var2Height <- plotData$Freq / plotData$marginVar1
  plotData$var1Center <- c(0, cumsum(plotData$marginVar1)[1:levVar1 -1]) +
    plotData$marginVar1 / 2

  ggplot(plotData, aes(var1Center, var2Height)) + 
    geom_bar(stat = "identity", aes(width = marginVar1, fill = var2), 
             col = "Black") +
    xlab(xlabel) + #added these three lines and deleted 1 geom_text- Jas
    ylab(ylabel) +
    ggtitle(title)
  }
#remove unknown values and only the vehicle types that represent majority 
mdf <- subset(df, !(df$P_SAFE %in% c("NN","QQ","UU")) & 
          (df$V_TYPE %in% c("01","05","06","07","08"))) 
#either a safety device was used or not - 01 - No Safety, all others 2 - Safety
mdf$P_SAFE <- ifelse(mdf$P_SAFE == "01","01","02") 

g1 <- ggMMplot(mdf$V_TYPE[mdf$M==1],mdf$P_SAFE[mdf$M==1], "Vehicle Type 
        (V_TYPE)", "Safety Device Used (P_SAFE)","Safety Device Used by Vehicle 
        Type\n(Deadly Collisions)")
g2 <- ggMMplot(mdf$V_TYPE[mdf$M==0],mdf$P_SAFE[mdf$M==0], "Vehicle Type 
        (V_TYPE)", "Safety Device Used (P_SAFE)", "Safety Device Used by Vehicle 
        Type\n(Non-Deadly Collisions)")
```

```{r, out.width='1000px', dpi=200}
g2 + theme(legend.position = "none",axis.text.x = element_text(size = 8, 
      angle = 0, vjust = 0.5, hjust=1),plot.title = element_text(lineheight=.8, 
      face="bold")) +
    annotate("text", label = "Safety Device Used", x=0.1 , hjust=0, y=0.5, 
             color= "black", size=10) +
    annotate("text", label = "Light Duty Vehicles (1)", x=0.0 , 
             hjust=0, y=1.05, color= "black", size=5) +
    annotate("text", label = "(5)", x=0.93 , hjust=0, y=1.05, 
             color= "black", size=3) +
    annotate("text", label = "(6)", x=0.95 , hjust=0, y=1.05, 
             color= "black", size=3) +
    annotate("text", label = "(7)", x=0.97 , hjust=0, y=1.05, 
             color= "black", size=3) +
    annotate("text", label = "(8)", x=0.99 , hjust=0, y=1.05, 
             color= "black", size=3) +
    scale_y_continuous(breaks = seq(0,1,0.1), labels = seq(0,1,0.1)) +
    scale_x_continuous(breaks = seq(0,1,0.1), labels = seq(0,1,0.1)) 
```

```{r, fig.width=8, fig.height=6}
g1 + theme(legend.position = "none",axis.text.x = element_text(size = 8, 
      angle = 0, vjust = 0.5, hjust=1),plot.title = element_text(lineheight=.8, 
      face="bold")) +
    annotate("text", label = "No Safety Device Used", x=0.1 , hjust=0, y=0.1, 
             color= "black",size=10) +
    annotate("text", label = "Safety Device Used", x=0.1 , hjust=0, y=0.5, 
             color= "black", size=10) +
    annotate("text", label = "Light Duty Vehicles (1)", x=0 , 
             hjust=0, y=1.05, color= "black", size=5) +
    annotate("text", label = "(5)", x=0.82 , hjust=0, y=1.05, 
             color= "black", size=5) +
    annotate("text", label = "(6)", x=0.85 , hjust=0, y=1.05, 
             color= "black", size=5) +
    annotate("text", label = "(7)", x=0.90 , hjust=0, y=1.05, 
             color= "black", size=5) +
    annotate("text", label = "(8)", x=0.95 , hjust=0, y=1.05, 
             color= "black", size=5) +
    scale_y_continuous(breaks = seq(0,1,0.1), labels = seq(0,1,0.1)) +
    scale_x_continuous(breaks = seq(0,1,0.1), labels = seq(0,1,0.1))  
```

#### We can see that light duty vehicles (1) represent the largest proportion of all collisions (including deadly collisions). A large proportion of deadly accidents also occured when no safety device (RED), such as a seatbelt, was used. This is consistent across vehicle types.

### Plot Three
```{r, echo=FALSE, Plot_Three, out.width='1000px', dpi=200}
#No need to manually rename all the labels, only the one that is interesting
collision_types <- levels(df$C_CONF)
#length(collision_types) #need to know the length so I can subset out UU & QQ
collision_types <- collision_types[1:18] #subset out UU & QQ
collision_types[which(collision_types == "31")] <- "31 - Head-on collision"
ggplot(aes(x=V_TYPE, y=C_CONF, color = factor(M,labels=c("Non-Deadly",
    "Deadly"))), data=subset(df,C_CONF != "UU" & C_CONF != "QQ")) + 
    geom_point(position="jitter") +
    ggtitle("Vehicle Type vs Collision Type") + 
    xlab("Vehicle Type\n(V_TYPE)") + 
    ylab("Collision Type\n(C_CONF)") +
    theme(legend.position = "top",axis.text.x = element_text(size = 8, 
      angle = 0, vjust = 0.5, hjust=1),plot.title = element_text(lineheight=.8, 
      face="bold")) +
    scale_x_discrete(labels = vehicle_types) + #vehicle types created earlier
    labs(color="Collision Severity") #rename legend
```

#### Head-on collisions (31) are the deadliest type of collisions for many vehicle types which is not a surprise, but its interesting to see that tractors in particular have a large proportion of deadly head-on collisions. You can also see that vast majority of accidents involve cars (or other light duty vehicles).

------

# Reflection

I wasn't able to satisfactorily answer the original question I posed as I need more statistical proof that the variables I suspected as being predictive of deadly crashes (due to the high proportions of these variables in deadly crashes as shown by the plots) are not due to just chance.

The features I would initially choose if I were building a machine learning model to predict deadly crashes are: *P_SEX*, *P_SAFE*, *V_TYPE* and *C_RCFG* ; from the plots they appear to differ the most between deadly/non-deadly crashes.

Once I complete the next machine learning course, I will be able to see which classification model would be appropriate. What I did learn through my exploratory analysis is the linear model would not be appropriate in this case.

Here are some valuable things I learned throughout this process:

### Reuse of Code

I learned that I would need to be very careful when subsetting data from a previous year in order to make it reusable. For example, I didn’t need to remove unknown variables for *C_SEV* because there were none for 2011, but that’s not necessarily true for other years. It is best to have the code in there assuming that the data has this missing data (even if it isn't missing in the particular dataset you are currently working on).

### Initial Bias

It's also tempting to stop exploring the data when you arrive at the answer that you originally expected from intuition. Its important to keep going as things may not be as they seem from just the initial look at the data or intuition.

### Summary vs Plots

I learned that a lot of the information about the data set could be gleaned just from the summary output (summmary(df)). However, many of the visualizations helped confirm these initial findings, but more importantly they helped in finding outliers and unusual relationships that were missed if you looked just at the summary. For example, it would not be able to see that older teenagers and elderly drivers are disproportionally responsible for deadly collisions.

### Speed of Execution

I learned that it's better to work with a smaller subset of the data in order to speed up execution of the code. For the pairwise plot, I used a sample of 10,000 records, but it still took several minutes to run this code even with less than 10% of the records. 

### Choosing the right plot

It's very important to choose the right type of plot or else you can get misleading results. For example, comparing weather and road surface by a simple scatterplot suffers from overplotting and it shows that wet roads and windy weather is associated with deadly crashes and all other combinations are associated with non-deadly crashes. But when you use jitter, the pattern is much more clear and you can see that this is not necessarily the case..

### Plotting in R Studio vs HTML Output

Finally, I want to discuss a frustration with plotting in R Studio vs the html output.
If you use the zoom plot option in R Studio, you expect the final output to be the same. 
However, I found that although I could change the plot output width and height, I found the plots didn't look as good as they did in R studio. 
