Analysis of Vehicle Collisions in Canada 
========================================================
### Jas Sohi 
#### March 29, 2015

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.
# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

###Set global options for all subsequent r code chunks 
library(knitr)
opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(ggplot2)
library(plyr)

#Global option to have a black and white theme for all plots
ggplot <- function(...) { ggplot2::ggplot(...) + theme_bw() }

#http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)
#multiplot function to plot multiple graphs on a single page

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist 
# (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

### About the Dataset

[National Collision Database (NCDB) - a database containing all police-reported motor vehicle collisions on public roads in Canada. Selected variables (data elements) relating to fatal and injury collisions for the collisions from 1999 to the most recent available data (2011).](http://open.canada.ca/data/en/dataset/1eb9eba7-71d1-4b30-9fb1-30cbdab7e63a)

### Goal

###*The underlying question I will explore in this analysis is what features (related to drivers, their vehicles, and the collision itself) are common in accidents that result in fatalities?*

### Data Wrangling

I chose to subset this data and only look at the most recent year (2011) due to the enormous size of the original dataset with over a million records. This new raw dataset contains just over 300,000 records.

Each record represents one collision report per person involved in a collision in Canada. So if there were 4 people involved in a collision there would be 4 separate records. 

```{r}
###DATA PREPARATION - This was done only once to subset the original da
###Set the working directory
# getwd()
# directory <- "C:/Users/user/version-control/Project-3-Data-Analysis-with-R"
# setwd(directory)
###Load up the original dataset
# df <- read.csv('NCDB_1999_to_2011.csv')
###Subset the dataset
# df2011 <- subset(df, C_YEAR == "2011")
# write.csv(df2011, "NCDB_2011.csv")
# rm(list=ls()) #removes all variables from the Global Environment 
```

```{r}
# Load the Data
filename <- "NCDB_2011.csv"
df <- read.csv(filename)
df2 <- df #Copy of the untouched dataframe for later comparison
```
### Number of records, variables 
```{r}
nrow(df) 
ncol(df)
```
### Variable names
```{r}
names(df)
```
[Data Dictionary (.doc)](http://data.tc.gc.ca/extracts/Data_Dictionary_NCDB_EN.doc)

# Univariate Analysis

### What is the structure of your dataset?

### Structure of the data
```{r}
str(df)
```

### Summary of the data
```{r}
summary(df)
```
There are three kinds of variables in this dataset: collision-level, vehicle-level, and person-level.

### COLLISIONS
From an initial look a the summary output, Friday (5) appears to be the weekday with the most number of collisions reported. The 4-5 PM Hour (16) is the most dangerous time for motorists. Most collisions involve 2 vehicles (*C_VEHS*). Rear-end collisions (21) are the most common types of accidents (*C_CONF*) and most accidents happen at an intersection (02) (*C_RCFG*). Nothing out of the ordinary so far. 
However, perhaps surprisingly, most accidents actually occur on clear and sunny days by a wide margin (1) (*C_WTHR*) and on dry,
normal (1) (*C_RSUR*) straight and level (1) 
(*C_RALN*) road surfaces. Finally, most collisions occur where no traffic controls (such as stop signs, police officers, reduced speed zone, etc.) are present.

### VEHICLES

Most collisions involved light duty vehicles such as passenger cars, vans, and light duty pickup trucks (01) (*V_TYPE*).
Most collisions involve more recent model vehicles, but there isn't a large variance between years (*V_YEAR*). 

### PEOPLE

```{r echo = FALSE}
#Divide the number of deadly collisions by the number of total collisions
percent_deadly <- round((nrow(subset(df,P_ISEV == 3))/nrow(df))*100,digits=2)
```
Men are involved in more collisions than women. Many people don't report their age (UU = Unknown) (*P_AGE*), but from the ones that do, most consist of young adults and teenagers (17-21 years old).About 2/3rds of the reported collisions (each record) are from drivers (11) themselves instead of passengers (*P_PSN*). Only ```r percent_deadly``` % of reported collisions involved a fatality (3), while the rest of the collisions were split mostly between injuries/no injuries (*P_ISEV*). In the vast majority of incidents, a safety device such as a seat belt or helmet (for motorcycles, bicyclists,
snowmobilers, and ATV riders) was worn (02) (*P_SAFE*).
This highlights the fact that there are different categories of people reporting these collisions, but the vast majority are vehicle drivers (1) and passengers (2) with pedestrians (3) a distant third (*P_USER*).  

# Univariate Plots Section

###Initial Age Plot
```{r}
qplot(x=P_AGE, data = df)
```
Looks like there is something unusual with this plot towards the right tail.
```{r}
#Let's see the frequency of each age
ages <- as.data.frame(table(df$P_AGE))
ages[ages$Var1 == "NN" | ages$Var1 == "UU",]
summary(df$P_AGE)[1:10]
```
### Further data wrangling
Looks like a lot of people did not report their age (UU) or was not applicable (NN) (e.g. “dummy" person record created for parked cars).
We'll go ahead and remove these records from the dataset we are looking and also clean up the gender variable to show only males and females. 
```{r}
df <- subset(df, P_AGE != "UU" & P_AGE != "NN" & P_SEX != "N" & P_SEX != "U")
#summary(df)
#275,270 records now
```
We are now looking at `r nrow(df)` records instead of just over three hundred thousand from the original data. We got rid of under 10% of the records and it shouldn't affect our ultimate analysis.

### Second Age Plot
```{r}
ggplot(aes(x=P_AGE), data = df) + geom_bar()
```
This chart now shows the distribution of ALL people involved in collisions and we can see that it is bimodal with one peak near later teenagers/young adults and another what looks like people in their 40s. Let's confirm that.

```{r}
#Needed to subset df & for later boxplot.
df$P_AGE_NUM <- as.numeric(df$P_AGE)
ggplot(aes(x=P_AGE), data = subset(df,P_AGE_NUM > 15)) + 
  geom_bar(binwidth = 1)
```
From this chart (age > 15), we can see more precisel that 45-50 is where the second peak is. But, as per the original goal of this analysis, we are really interested in the age of *Drivers* only.

###Driver's Age

```{r}
#P_USER is the type of user who reports the collision, 
#can be passenger, pedestrian, motorcyclists, etc.
ggplot(aes(x=P_AGE), data = subset(df,P_USER == 1)) + #1 for Driver
  geom_bar(binwidth = 1)
```
We do see a similar distribution when only considering Drivers only. We do see now that the left tail of the distribution is flat as we don't expect children to be driving (at least in Canada)!

### Gender

Does this age distribution vary across genders?
```{r}
ggplot(aes(x=P_AGE), data = subset(df, P_USER == 1)) + geom_bar() +
  facet_wrap(~P_SEX)
qplot(x = P_SEX, y = P_AGE_NUM, data = df, geom = "boxplot")
```
Not quite, the shape of both age distributions is very similar across sexes.

###Gender of the Driver

Are more men or women behind the wheel before an accident?

```{r}
#http://stackoverflow.com/questions/20600900/r-faceted-bar-chart-with-percentages-labels-independent-for-each-plot
ggplot(aes(x=P_SEX),data = subset(df,P_USER == 1)) + geom_bar() + 
  stat_bin(geom = "text", #show percentages on top of the bars
           aes(label = paste(round((..count..)/sum(..count..)*100), "%")),
           vjust = -1, color = "grey30", size = 5) +
           coord_cartesian(ylim=c(0,125000)) #leave room for the percent labels
```
[Contrary to this reddit joke](https://www.reddit.com/r/Jokes/comments/2zqnwx/fact_a_lot_of_women_turn_into_good_drivers/), in bad taste I may add, men are actually involved in more collisions than women overall.

### Deadly Crashes

Is this difference between the sexes still present when we look at deadly crashes only?

```{r}
#C_SEV Severity of the crash - At least one fatality(1) or non-fatal injuries(2) 
ggplot(aes(x=P_SEX),data = subset(df,P_USER == 1 & C_SEV == 1)) + geom_bar() +  
  stat_bin(geom = "text", #show percentages on top of the bars
           aes(label = paste(round((..count..)/sum(..count..)*100), "%")),
           vjust = -1, color = "grey30", size = 5) +
           coord_cartesian(ylim=c(0,2250)) #leave room for the percent labels
```

Yes, and the contrast is much more pronounced. There are relatively a few deaths per year (thank god!), but in 3 out of every 4 collisions that do involve a death, a man was driving the vehicle.

### Time

Let's shift gears (no pun intended) and look at collisions in regards to time.

### What day of the week has the most accidents?

```{r}
ggplot(aes(x=C_WDAY), data = subset(df,C_WDAY != "U")) + geom_bar()
```

Looks like Friday (5) has the most accidents (an increasing trend from Monday - perhaps people are in a rush to celebrate the weekend!) and Sunday has the least amount of accidents (this could be attributed to the fact that very few people work on Sundays so there might be less vehicles commuting to work overall).

```{r}
g1 <- ggplot(aes(x=C_WDAY), data = subset(df,C_WDAY != "U" & C_SEV == 1)) + 
  geom_bar() + ggtitle("Deadly Collisions")
g2 <- ggplot(aes(x=C_WDAY), data = subset(df,C_WDAY != "U" & C_SEV == 2)) + 
  geom_bar() + ggtitle ("Non-Deadly Collsions")
multiplot(g1, g2, cols=2)
```

Looks like deadly collisions happen the most on Saturdays (6) as opposed to Fridays (5) for non-deadly crashes. 

### What time of day do most collisions occur?

```{r}
ggplot(aes(x=C_HOUR), data = df) + geom_bar()
```

Not surprisingly, most collisions occur during 4:00-5:00 PM (not surprisingly many people are heading home after work during this hour).

```{r}
g1 <- ggplot(aes(x=C_HOUR), data = subset(df,C_SEV == 1)) + geom_bar() + 
  ggtitle("Deadly Collisions")
g2 <- ggplot(aes(x=C_HOUR), data = subset(df,C_SEV == 2)) + geom_bar() + 
  ggtitle ("Non-Deadly Collsions")
multiplot(g1, g2, cols=2)
```

### How about if we look at time of day across the different days?
```{r}
ggplot(aes(x=C_HOUR), data = subset(df,C_WDAY != "U" & C_HOUR != "UU")) + 
  geom_bar() +
  facet_wrap(~C_WDAY)
```
Looks like distribution across the weekdays is bimodal throughout the non-sleeping hours (Midnight to 7 AM   ), while the distribution on the weekend is almost identical for saturday and sunday in shape being nearly unimodal for the non-sleeping hours.

### What month has the most accidents?

```{r}
ggplot(aes(x=C_MNTH), data = subset(df, C_MNTH != "UU")) + geom_bar()
```
Looks like January and the summer months have the most number of reported collisions (could be that more people are driving in the summer for the holidays).

### Does this change when we only look at drivers?
```{r}
ggplot(aes(x=C_MNTH), data = subset(df, C_MNTH != "UU" & P_USER == 1)) + 
  geom_bar()
```
Not really, the absolute counts are less, but the distribution is quite similar.

### Types of People Reporting Collisions

Speaking of drivers, we haven't yet discussed how much of this dataset consists of collision reports from drivers as opposed to passengers, pedestrians, etc.

```{r}
ggplot(aes(x=P_USER), data = subset(df, P_USER != "U")) + geom_bar() +
  stat_bin(geom = "text", #show percentages on top of the bars
           aes(label = paste(round((..count..)/sum(..count..)*100), "%")),
           vjust = -1, color = "grey30", size = 4) +
           coord_cartesian(ylim=c(0,200000)) #leave room for the percente labels
```
The vast majority of this dataset comes from reports from drivers (1) and passengers (2).

### How's the weather like on when there are accidents?
```{r}
ggplot(aes(x=C_WTHR), data = df) + geom_bar()
```
Perhaps counterintuitively, most accidents occur on sunny days. At first, you might have thought that more accidents occur in bad weather.

### Let's compare the weather for motorcyclists vs drivers
```{r}
g1 <- ggplot(aes(x=C_WTHR), data = subset(df, P_USER == 5)) + #5 - Motorcyclists
  geom_bar()
  #More accidents in rain and snow
g2 <- ggplot(aes(x=C_WTHR), data = subset(df, P_USER == 1)) + #1 - Drivers
  geom_bar()
multiplot(g1, g2, cols=2)  
```
A larger proportion of the accidents that motorcyclists (left plot) are involved in occur on sunny days (1) than compared with motor vehicle drivers (right plot). (this could be due to the fact that a lot of motorcyclists ride only in the spring and summer months with better weather conditions) 

### Number of vehicles involved in a collision
```{r}
qplot(x=C_VEHS, data = subset(df, P_USER == 1)) 
```
Most collisions involve, not surprisingly, 2 vehicles.

### Vehicle Age
```{r}
qplot(x=V_YEAR, data = subset(df, P_USER == 1)) 
```

```{r}
g1 <- ggplot(aes(x=V_YEAR), data = subset(df, P_USER == 1 & C_SEV == 1)) +
  geom_bar() +
  ggtitle("Deadly Collisions")
g2 <- ggplot(aes(x=V_YEAR), data = subset(df, P_USER == 1 & C_SEV == 2)) +
  geom_bar() +
  ggtitle("Non-Deadly Collisions")
multiplot(g1,g2, ncol = 2)
```

Let's cleanup the plot a bit and only look at vehicles manufactured after 1980.

```{r}
#Need to create a new numeric variable to be able to subset the data below
df$V_YEAR_NUM <- as.numeric(as.character(df$V_YEAR))
g1 <- ggplot(aes(x=V_YEAR), data = subset(df, P_USER == 1 & C_SEV == 1 &
  V_YEAR != "NNNN" & V_YEAR != "UUUU" & df$V_YEAR_NUM >= 1980)) +
  geom_bar() +
  ggtitle("Deadly Collisions")
g2 <- ggplot(aes(x=V_YEAR), data = subset(df, P_USER == 1 & C_SEV == 2 & 
  V_YEAR_NUM >= 1980 & V_YEAR != "NNNN" & V_YEAR != "UUUU")) +
  geom_bar() +
  ggtitle("Non-Deadly Collisions")
multiplot(g1,g2, ncol = 2)
```

```{r}
qplot( geom = "boxplot", x= factor(C_SEV), y = V_YEAR_NUM, data = df)
```

Doesn't seem to be any clear trend other than most vehicles involved in accidents are from most recent models (it could simply be there are more newer vehicles on the road). Looks like a dead end here (pun intended!).

###Vehicle Type

Okay, so now the type of vehicle you are driving has to be a factor in determining whether deaths are involved in collisions right? Let's see.

```{r echo=FALSE, Summary}
qplot(x=V_TYPE, data = df)
```

Looks like the vast majority of vehicles involved in accidents are Light Duty Vehicles (1).

```{r}
g1 <- ggplot(aes(x=V_TYPE), data = subset(df,P_USER == 1 & C_SEV == 1)) +
  geom_bar() +
  ggtitle("Deadly Collisions")
g2 <- ggplot(aes(x=V_TYPE), data = subset(df,P_USER == 1 & C_SEV == 2)) +
  geom_bar() +
  ggtitle("Non-Deadly Collisions")
multiplot(g1,g2, ncol = 2)
```
Now remember we are now looking at drivers only; we see that a larger proportion of Trucks (07) & Road Tractors (08) are involved in deadly crashes. This may be one variable to help predict deadly crashes.

#Univariate Analysis (continued)

### What is/are the main feature(s) of interest in your dataset?

The main features of interest in the data set are *IS_SEV* (the severity of collisions - injury, no injury, or fatality) and *P_AGE* (the age of person involved in collision). I suspected that driver's age and some other combination of variables could help predict deadly collisions.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

In addition to the age of the driver, I originally thought that vehicle age (*V_YEAR)*, vehicle type (*V_TYPE*), and gender (*P_SEX*) would be a larger factor in determining the fatality in collisions. The plots support this initial hunch except for vehicle age which doesn't seem to vary between deadly and non-deadly collisions. I will look further at age in the bivariate analysis to see if their is a difference between deadly and non-deadly collisions.

### Did you create any new variables from existing variables in the dataset?

Yes, I did create new variables just so that the data could be subset since the original variables were factors (categorical) and could not be subsetted otherwise (P_AGE_NUM & V_YEAR_NUM). I didn't want to directly convert the original columns in case I needed the factor variables at a later time.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

Yes, I did adjust the form of the data to show only the most recent year of data as there were millions of total records, but I was able to subset it to 2011 and bring it down to 300,000 records and then I removed data that had unknown or missing gender values as I suspected that the sex of the driver in fatal collisions could be skewed towards one sex (as confirmed by a histogram plot).

# Bivariate Plots Section
```{r}
#http://stackoverflow.com/questions/3571909/r-how-to-calculate-correlation-cor-for-only-a-subset-of-columns
# Make the correlation table
remove <- names(df) %in% c("C_YEAR") #We are not selecting the year when producing this correlation table as it produces NA values

rdf <- df[!remove]
ctab <- cor(rdf[sapply(rdf, is.numeric)])
round(ctab, 2)
```

Let's see what other variables can be considered numeric.

```{r echo=TRUE}
df$C_MNTH_NUM <- as.numeric(df$C_MNTH); 
df$C_WDAY_NUM <- as.numeric(df$C_WDAY)
df$C_HOUR_NUM <- as.numeric(df$C_HOUR); 
df$C_VEHS_NUM <- as.numeric(df$C_VEHS) #number of vehicles
df$C_MNTH_NUM <- as.numeric(df$C_MNTH); 
df$C_WTHR_NUM <- as.numeric(df$C_WTHR) #higher value = worse weather
df$C_RSUR <- as.numeric(df$C_WTHR) #higher value = worse driving conditions
```

I now realized I had been looking at the collision level deaths (C_ISEV) when to answer my question I should really be looking at the person level (P_ISEV). I should be predicting if a driver is involved in a crash, what factor predicts whether he or she will survive? 

I will actually create a new variable - Driver Mortality (DM) - 0 is the driver has survived (No Injury or Injurty) or has died - 1? This is what I will try to predict.

Let's look back at P_ISEV to see if there are alot of unknown values (U) or not applicable (N - e.g. dummy cars)?

```{r}
summary(df$P_ISEV)
```

I'm going to go ahead and remove records with these two values and create the new (DM) variable as mentioned before.

```{r}
#275,270 records before
df3 <- df  #create a copy in case need to revert due to unintended changes
#X - jurisdiction does not provide info - is not found in this 2011
#data but may be if we look at other years
df <- subset(df, P_ISEV != "U" & P_ISEV != "N" & P_ISEV != "X")  
#http://www.statmethods.net/management/variables.html
df$DM <- ifelse(df$P_ISEV == 3, 1, 0) 
table(df$DM)
```
*1* means that that record for that person was deceased immediately after the collision or went to the hospital and died as a result of his/her injuries. *0* is the person survived.

That removed about 10,000 records. We will make sure to make this same transformation to the test data we will test our model on. We will now subset the data to only drivers.

```{r}
df4 <- df
#264,459 records before
df <- subset(df, P_USER == 1)
nrow(df)
```
Now we have reduced the records by just under 100,000 but we are looking at only the drivers and we are ready to apply our model. Just before we do let's have another look at the correlation matrix now.

```{r}
ctab <- cor(df[sapply(df, is.numeric)])
round(ctab, 2)
```
Definately, doesn't seem to be any linear relationsips here.

Let's go back to draw some bivariate plots.

#Bivariate Plots
```{r}
#https://tgmstat.wordpress.com/2013/11/13/plot-matrix-with-the-r-package-ggally/
library(GGally)
#P_PSN,P_USER don't need to be looked at as they both indicate drivers only
remove_columns <- names(df) %in% 
  c("X","C_YEAR","C_MNTH", "C_HOUR", "V_ID","V_YEAR","P_ID", "P_AGE", "P_ISEV",
    "C_SEV", "P_USER", "C_WTHR", "C_VEHS", "C_CONF", "C_RCFG","C_TRAF","P_SAFE",
    "P_PSN", "C_WDAY") 
sampledf <- df[!remove_columns]
#Rename the variables for better visibility on the plot
names(sampledf) <- c("RS","RL","VT","SEX","AGE","VY","M","D","H","VN","WT","DM")
library(GGally)
ggpairs(data=sampledf[1:10000, c(3:12)], # data.frame with variables,
        title="pairwise plots") # title of the plot
```


###Drivers in Newer Vehicles or Older Vehicles?
```{r}
ggplot(aes(x=factor(DM), y=V_YEAR_NUM), data=subset(df,V_YEAR != "UUUU" & 
      V_YEAR != "NNNN")) + geom_boxplot()
```
Looks like newer cars are involved in most accidents whether deadly(1) or not (0).

### Vehicle Types

Let's look at the age distribution of drivers of different types of vehicles.
```{r}
ggplot(aes(x=V_TYPE,y=P_AGE_NUM), data = df) + 
  geom_boxplot() +
  scale_x_discrete(labels=c("CAR","VAN","SMALL TRUCK","TRUCK","TRACTOR","SCHOOL BUS", "SMALL SCHOOL BUS", "CITY BUS", "RV", "FIRETRUCK","STREET CAR")) 
```

Nothing out of the ordinary, we can see that school bus drivers and RV drivers skew towards older drivers.

```{r}
qplot(x=C_TRAF, geom="bar", fill=factor(DM), data = df)
```

Initially it looks like most deaths occur where no safety controls are present (18)

```{r}
ggplot(df, aes(x = C_TRAF, fill = factor(DM))) + geom_bar(position = "fill")
```

But looking at the proportions, it looks as though the largest proportion of deaths actually occur where warning signs (5) are present.

###Age of Deadly Drivers
```{r}
library(vcd)
#http://cran.us.r-project.org/web/packages/vcdExtra/vignettes/vcd-tutorial.pdf
cdplot(factor(DM) ~ P_AGE, data = df)
#with(Arthritis, rug(jitter(Age), col="white", quiet=TRUE))
```
This cumulative density plot shows teenage drivers and elderly drivers are overrepresented in collision deaths.

###Underage Drivers
```{r}
#http://en.wikipedia.org/wiki/Driver%27s_licence_in_Canada
underage <- nrow(df[df$P_AGE_NUM < 14,])
underage_deaths <- df[df$P_AGE_NUM < 14 & df$DM == 1,]
```
Why are there some young drivers? The lowest legal age for obtaining anytype of driver's license anywhere in Canada is 14 (Alberta). Looks like there are `r underage` cases of underage drivers in 2011 and only `r nrow(underage_deaths)` driver, aged `r underage_deaths$P_AGE`, resulted in a deadly collision (see chart below). 

```{r}
ggplot(aes(x=P_AGE, fill= factor(DM)), data = df) + geom_bar(position = "fill")
```


```{r}
# mosaic(df, gp = shading_max, split_vertical = TRUE, main="Arthritis: [P_USER] [P_SEX]")
# mosaic(~ C_WDAY + DM + C_HOUR, data=df,
# expected=~C_WDAY:C_HOUR + C_WDAY:DM + C_HOUR:DM,
# legend=FALSE, gp=shading_Friendly)

```

Maybe the configuration of the crash has an effect on how deadly a crash is?

```{r}
ggplot(aes(x=C_CONF, fill= factor(DM)), data = df) + geom_bar(position = "fill")
```

Head on collisions (31)  make up a huge proportion of deadly crashes. 

```{r}
ggplot(aes(x=C_RCFG, fill= factor(DM)), data = df) + geom_bar(position = "fill")
```

Railroad crossings (4) and passing lanes (7)  are the type of road configurations
with the most deadly crashes.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?


### What was the strongest relationship you found?



# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
# qplot(x=P_AGE, y=V_YEAR, data=subset(df,V_YEAR != "UUUU" & V_YEAR != "NNNN"), color = P_SEX)
```

```{r}
# qplot(x=P_AGE, y=V_YEAR, data=subset(df,V_YEAR != "UUUU" & V_YEAR != "NNNN"), color = factor(DM))
```

```{r}
# totalDeaths <- nrow(df[df$DM == 1,])
# total <- nrow(df)
# totalDeaths/total
# Deaths <- df[df$DM == 1,]
```
#ggplot((x=age))
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------


# Final Plots and Summary

### Plot One
```{r echo = FALSE, eval=FALSE}
# To help see what limits to set on next plot
boxplot(df$P_AGE_NUM) 
summary(df$P_AGE_NUM)
```

```{r out.width='1000px', dpi=200}
#http://stackoverflow.com/questions/23709060/change-r-markdown-plot-width

#how may deaths for the most deadly age (19) and the second peak
age_most <- 19
age_old <- 46 #2nd peak
age_initial <- 14 #lowest age on the bar chart
most_deaths <- length(which(df$P_AGE_NUM == age_most))
older_deaths <- length(which(df$P_AGE_NUM == age_old))
#I will choose 14 as the minimum since that is the lowest legal age
ggplot(aes(x=P_AGE), data = subset(df,P_AGE_NUM >= age_initial & P_AGE_NUM <= 90)) + 
  geom_bar(binwidth = 1, colour = "darkgreen", fill = "white") +
  ggtitle("Histogram of Driver's Age (All Collisions)") + 
  xlab("Driver's Age") + 
  ylab("Count of Drivers") +
  #http://stackoverflow.com/questions/11019692/place-annotation-at-the-top-of-a-series-of-histograms-in-ggplot2-using-a-for-loo
  annotate("text", label = as.character(most_deaths), x= age_most - age_initial, 
    hjust = 0, y = most_deaths + 100, color = "black") + #Show the exact number
  annotate("text", label = as.character(older_deaths), x= age_old - age_initial, 
    hjust = 0, y = older_deaths + 100, color = "black") + #Show the exact number
   #added 100 to the y-axis so there would be no overlap between text and plot
#http://stackoverflow.com/questions/1330989/rotating-and-spacing-axis-labels-in-ggplot2
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))
```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection
