========================================================
Analysis of Vehicle Collisions in Canada by Jas Sohi

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.
# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.
library(knitr)
opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(ggplot2)
library(leaflet) #For interactive maps
library(DT) #Datatables Javascript
library(maps)
library(plyr)

#Global Option to have a black and white theme for all plots
ggplot <- function(...) { ggplot2::ggplot(...) + theme_bw() }

#http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)
#multiplot function to plot multiple graphs on a single page

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

```

ABOUT THE DATASET

[National Collision Database (NCDB) - a database containing all police-reported motor vehicle collisions on public roads in Canada. Selected variables (data elements) relating to fatal and injury collisions for the collisions from 1999 to the most recent available data.](http://open.canada.ca/data/en/dataset/1eb9eba7-71d1-4b30-9fb1-30cbdab7e63a)

###Goal

The underlying question I will explore in this analysis is what features are common in drivers for accidents that result in fatalities?

###Data Wrangling

I chose to subset this data and only look at the most recent year (2011) due to the enormous size of this dataset with over a million records. This subsetted dataset contains just over 300,000 records.

Each record represents one collision report per person involved in a collision in Canada. So if there were 4 people involved in a collision there would be 4 separate records. 

```{r}
###DATA PREPARATION
###Set the working directory
# getwd()
# directory <- "C:/Users/user/version-control/Project-3-Data-Analysis-with-R"
# setwd(directory)
###Load up the original dataset
# df <- read.csv('NCDB_1999_to_2011.csv')
###Subset the dataset
# df2011 <- subset(df, C_YEAR == "2011")
# write.csv(df2011, "NCDB_2011.csv")
# rm(list=ls()) #removes all variables from the Global Environment 
```

```{r}
# Load the Data
filename <- "NCDB_2011.csv"
df <- read.csv(filename)
df2 <- df #Copy of the untouched dataframe for later comparison
```
###Number of records, variables 
```{r}
nrow(df) 
ncol(df)
```
###Variable names
```{r}
names(df)
```
[Data Dictionary (.doc)](http://data.tc.gc.ca/extracts/Data_Dictionary_NCDB_EN.doc)

###Structure of the data
```{r}
str(df)
```

###Summary of the data
```{r}
summary(df)
```
There are three kinds of variables in this dataset: collision-level, vehicle-level, and person-level.

###COLLISIONS
From an initial look a the summary output, Friday (5) appears to be the weekday with the most number of collisions reported. The 4-5 PM Hour (16) is the most dangerous time for motorists. Most collisions involve 2 vehicles (*C_VEHS*).Rear-end collisions (21) are the most common types of accidents (*C_CONF*) and most accidents happen at an intersection (02) (*C_RCFG*). Nothing out of the ordinary so far. 
However, perhaps surprisingly most accidents actually occur on clear and sunny days by a wide margin (1) (*C_WTHR*) and on dry,
normal (1) (*C_RSUR*) straight and level (1) 
(*C_RALN*) road surfaces. Finally, most collisions occur where no traffic controls (such as stop signs, police officers, reduced speed zone, etc.) are present.

###VEHICLES

Most collisions involved light duty vehicles such as passenger cars, vans, and light duty pickup trucks (01) (*V_TYPE*).
Most collisions involve more recent model vehicles, but there doesn't seem to be a large variance between years (*V_YEAR*). 

###PEOPLE

Men are involved in more collisions than women. Many people don't seem to report their age (UU = Unknown) (*P_AGE*), but from the ones that do, most consist of young adults and teenagers (17-21 years old).

About 2/3rds of the reported collisions (each record) are from drivers (11) themselves instead of passengers (*P_PSN*). Only
```r round((nrow(subset(df,P_ISEV == 3))/nrow(df))*100,digits=2) ``` % of reported collisions involved a fatality (3), while the rest of the collisions were split mostly between injuries/no injuries (*P_ISEV*). In the vast majority of incidents, a safety device such as a seat belt or helmet (for motorcycles, bicyclists,
snowmobilers, and ATV riders) was worn (02) (*P_SAFE*).
This highlights the fact that there are different categories of people reporting these collisions, but the vast majority are vehicle drivers (1) and passengers (2) with pedestrians (3) a distant third (*P_USER*).  

###Initial Age Plot
```{r}
qplot(x=P_AGE, data = df)
```
Looks like there is something unusual with this plot towards the right.
```{r}
#Let's see the frequency of each age
ages <- as.data.frame(table(df$P_AGE))
ages[ages$Var1 == "NN" | ages$Var1 == "UU",]
summary(df$P_AGE)[1:10]
```
###Further data wrangling
Looks like a lot of people did not report their age (UU) or was not applicable (NN) (e.g. “dummy" person record created for parked cars).
We'll go ahead and remove these records from the dataset we are looking and also clean up the gender variable to show only males and females. 
```{r}
df <- subset(df, P_AGE != "UU" & P_AGE != "NN" & P_SEX != "N" & P_SEX != "U")
nrow(df)
summary(df)
```
We are now looking at 275,270 records instead of just over three hundred thousand from the original data. We got rid of under 10% of the records and it shouldn't affect our ultimate analysis.

###Age
```{r}
ggplot(aes(x=P_AGE), data = df) + geom_bar()
```
This chart now shows the distribution of ALL people involved in collisions and we can see that it is bimodal with one peak for teenages and another what looks like around the 40s. Let's confirm that.

```{r}
df$P_AGE_NUM <- as.numeric(df$P_AGE)
ggplot(aes(x=P_AGE), data = subset(df,P_AGE_NUM > 20)) + 
  geom_bar(binwidth = 1)

```
From this chart (age > 20) it looks like it is about 45-50 where the second peak is.

But we are really interested in Drivers only in terms of their age.

###Driver's Age

```{r}
df$P_AGE_NUM <- as.numeric(df$P_AGE)
#P_USER is the type of user who reports the collision, can be passenger, pedestrian, etc (Driver - includes motorcycle drivers/riders) 
ggplot(aes(x=P_AGE), data = subset(df,P_USER == 1)) + #1 for Driver
  geom_bar(binwidth = 1)
```
We do see a similar distribution when only considering Drivers. We do see now that the left end of the distribution is flat as we don't expect children to be driving (at least in Canada)!

###Gender

Does this age distribution vary between genders?
```{r}
ggplot(aes(x=P_AGE), data = subset(df, P_USER == 1)) + geom_bar() +
  facet_wrap(~P_SEX)
```
Not quite, although the 1st peak is much higher for Male Drivers and the 2nd peak is more prominent in the Male Driver distribution, the overall shape of both distributions are very similar.

###Gender of the Driver

Are more men or women behind the wheel before an accident?

```{r}
#http://stackoverflow.com/questions/20600900/r-faceted-bar-chart-with-percentages-labels-independent-for-each-plot
ggplot(aes(x=P_SEX),data = subset(df,P_USER == 1)) + geom_bar() + 
  stat_bin(geom = "text", #show percentages on top of the bars
           aes(label = paste(round((..count..)/sum(..count..)*100), "%")),
           vjust = -1, color = "grey30", size = 5) +
           coord_cartesian(ylim=c(0,125000)) #leave room for the bar labels by increasing max
```
[Contrary to this reddit joke](https://www.reddit.com/r/Jokes/comments/2zqnwx/fact_a_lot_of_women_turn_into_good_drivers/) (in bad taste), men may actually be involved in more collisions than women overall.

###Deadly Crashes

Is there a difference between the sexes when we look at deadly crashes only?

```{r}
#CSEV --> Severity of the the crash - Either at least one fatality(1) or non-fatal injuries(2) 
ggplot(aes(x=P_SEX),data = subset(df,P_USER == 1 & C_SEV == 1)) + geom_bar() + #1 
  stat_bin(geom = "text", #show percentages on top of the bars
           aes(label = paste(round((..count..)/sum(..count..)*100), "%")),
           vjust = -1, color = "grey30", size = 5) +
           coord_cartesian(ylim=c(0,2250)) #leave room for the bar labels by increasing max
```

Yes, there are relatively few deaths (thank god!), but in 3 out of every 4 collisions that involve a death, a man was driving the vehicle.

###Time

Let's shift gears (no pun intended) and look at collisions in regards to time.

###What day of the week has the most accidents?

```{r}
ggplot(aes(x=C_WDAY), data = subset(df,C_WDAY != "U")) + geom_bar()
```

Looks like Friday (5) has the most accidents (an increasing trend from Monday) and Sunday has the least amount of accidents (this could be attributed to the fact that very few people work on Sundays so there might be less vehicles commuting to work).

```{r}
g1 <- ggplot(aes(x=C_WDAY), data = subset(df,C_WDAY != "U" & C_SEV == 1)) + geom_bar() + ggtitle("Deadly Collisions")
g2 <- ggplot(aes(x=C_WDAY), data = subset(df,C_WDAY != "U" & C_SEV == 2)) + geom_bar() + ggtitle ("Non-Deadly Collsions")
   #leave room for the bar labels by increasing max
multiplot(g1, g2, cols=2)
```

Looks like deadly collisions happen the most on Saturdays as opposed to friday for non-deadly crashed.

###What time of day do most collisions occur?

```{r}
ggplot(aes(x=C_HOUR), data = df) + geom_bar()
```
Not surprisingly, most collisions occur during 4:00-5:00 PM (many people are heading home after work during this hour).

#How about if we look at time of day across the different days?
```{r}
ggplot(aes(x=C_HOUR), data = subset(df,C_WDAY != "U" & C_HOUR != "UU")) + geom_bar() +
  facet_wrap(~C_WDAY)
```
Looks like the weekend, the distribution is almost identical in shape and the pake is much wider.

###What month has the most accidents?

```{r}
ggplot(aes(x=C_MNTH), data = subset(df, C_MNTH != "UU")) + geom_bar()
```
Looks like January and the summer months have the most number of reported collisions (could be that more people are driving in the summer for the holidays).

###Does this change when we only look at drivers?
```{r}
ggplot(aes(x=C_MNTH), data = subset(df, C_MNTH != "UU" & P_USER == 1)) + geom_bar()
```
Not really, the absolute counts are less, but the distribution is quite similar.

#Types of People Reporting Collisions

Speaking of drivers, we haven't yet discussed how much of this dataset consists of collision reports from drivers as opposed to passengers, pedestrians, etc.

```{r}
ggplot(aes(x=P_USER), data = subset(df, P_USER != "U")) + geom_bar() +
  stat_bin(geom = "text", #show percentages on top of the bars
           aes(label = paste(round((..count..)/sum(..count..)*100), "%")),
           vjust = -1, color = "grey30", size = 4) +
           coord_cartesian(ylim=c(0,200000)) #leave room for the bar labels by increasing max

```
The vast majority of this dataset comes from reports from drivers (1) and passengers(2).



#How's the weather like on when there are accidents?
#On sunny days, perhaps its just because more motorcyclists avoid bad weather.
qplot(x=C_WTHR, data = subset(df, P_USER == 5)) #5 is Motorcyclists
#More accidents in rain and snow
qplot(x=C_WTHR, data = subset(df, P_USER == 1)) + #5 is Driver
  theme(panel.background = element_rect(fill = 'white'))

#What day of the week has the most accidents?
qplot(x=C_WDAY, data = subset(df, P_USER == 1)) + 
  theme(panel.background = element_rect(fill = 'white'))
#Friday - Perhaps people are in a rush to celebrate the weekend!
#On Sundays the least amount of accidents, perhaps less people are on the road commuting to work.
#How about time of the day?
qplot(x=C_HOUR, data = subset(df, P_USER == 1)) +
  theme(panel.background = element_rect(fill = 'white'))
#How about day and hour?
qplot(x=C_HOUR, data = subset(df, P_USER == 1)) +
  facet_wrap(~C_WDAY) +
  theme(panel.background = element_rect(fill = 'white'))
#Number of vehicles involved in a collision
qplot(x=C_VEHS, data = subset(df, P_USER == 1)) +
  theme(panel.background = element_rect(fill = 'white'))
```

```{r echo=FALSE, Summary}
# Summarise the Data
#What month has the most accidents? Mosaic plot 
```


# Univariate Plots Section
```{r echo=FALSE, Univariate_Plots}

```

# Univariate Analysis

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

The main features of interest in the data set are *IS_SEV* (the severity of collisions - injury, no injury, or fatality) and *P_AGE* (the age of person involved in collision). I suspect age and some other combination of variables can be helpful in determining what are common variables in deadly collisions.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

In addition to the age of the driver, I think vehicle age (*V_YEAR)*, vehicle type (*V_TYPE*), and gender (*P_SEX*) would be predictive of fatality in collisions. Intuitively, I would expect young teenage males with older vehicles and 


### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?



# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

### What was the strongest relationship you found?




# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection
